Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-05-21 Multiresolution_framework/ltx/manuscript.tex
added Viktor comments

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}                               
\usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}

\newcommand{\dean}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
\newcommand{\vik}[1]{\textcolor{red}{#1}}

@@

\begin{abstract}
%\boldmath
\parham{Neural fields are spatially continuous state variables described by integro-differential equations, which are well suited to describe the spatiotemporal evolution of cortical activations on multiple scales. Here we} develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximization (EM) algorithm
\end{keyword}

@@

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  Electroencephalography (EEG) and Magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (the pulse to wave function) and the activation function (wave to pulse function) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials in microscopic single neural models, such as integrate and fire, or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter, can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) to mean firing rates of neural masses. The key factor linking the key functions of the model to particular spatial scales is geometry of the connectivity, hence the ability to infer a multi-resolution connectivity kernel will enable to the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. Large-scale neuroinformatic efforts such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}) are on the way to enable patient-specific full brain modeling. This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data.

The first work (to the authors best knowledge) describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\citep{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme dubbed Dynamic Causal Modeling (DCM)~\citep{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \citep{Galka2008,schiff2008kalman,Daunizeau2009, Pinotsis2012}. Most recently, a framework was developed where a finite element model of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\citep{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously and completely.

@@

\caption{{\bf Examples of cardinal B-spline functions}. Piecewise polynomial functions (colored lines) are joined at the break points (black dots). (\textbf{A}) B-spline function of order 2 (linear B-spline). (\textbf{B}) B-spline function of order 8.}

@@

The $m^{th}$  order cardinal B-spline function is defined by the recurrence relation \citep{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \citet{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third degree polynomials added together at the joining points, with an explicit expression given by
\begin{align}
3!N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
One important feature of the B-spline functions is the \emph{partition of unity} property, meaning that the unity can be expressed as a linear sum of B-splines, i.e
\begin{equation}
	\sum_{l}N_m(r+l)\equiv1.
	\end{equation}
This is important as the underlying field can be reconstructed when it is saturated without introducing numerical errors which is not the case for example when Gaussian basis functions are used.

@@

 		\caption{{\bf The scaling and the wavelet function decomposition}. (\textbf{A}) The scaling function (blue) at $j=0$ is obtained using 625 scaling functions (black) at $j=4$. (\textbf{B}) The wavelet function (blue) at $j=0$ is obtained using 1250 scaling functions (black) at $j=4$.}

@@

It is important to point out there is no doubt that there will always be a mismatch between the model and actual cortical tissue. Nevertheless, the simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient-specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics, such as epilepsy. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling, where the dynamics and connectivity structure differs at different spatial scales~\citep{Jirsa2002,Jirsa2004,Qubbaj2009,Breakspear2005,Schultze-Kraft2011}.

@@

% In order to apply the framework to real data some assumptions must be made. A critical assumption is that the model provides an apt description of the cortical dynamics.
 \subsection{Modification to higher dimensions}
%To demonstrate MRAIDE framework, an assumption was made where the firing rate behaves linearly. The implementation of the framework in the nonlinear case can be addressed as future work. 
%Another future avenue would be the modification of the estimation framework to perform efficiently at higher dimensions.
 The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to  extract fine features of the field at vertical, horizontal and diagonal orientations. The formulations provided in this paper can be easily extended to two dimensions. However, the computational load of the estimation algorithm will be significant. It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model.
\subsection{Extensions to the framework} 
The authors acknowledge that there is, and will always be, a discrepancy between the model and cortex. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.

Additional extension for future work include extending the model to capture a second order synaptic response kernel, The second-order postsynaptic kernel proposed in \citet{VanRotterdam1982} leads to dynamics that are more reminiscent to real postsynaptic potentials. Another extension of this framework is the inclusion of finite distant-dependent action potential propagation velocity into the parametric neural field equations. In small networks the delays via propagation can be neglected. However, delays must be accounted for in large scale networks. For instance, it is shown by \citet{Ghosh2008} that inclusion of time delays arising from propagation along connecting fibres are essential for the fluctuations in resting state networks (see \citet{Deco2010} for review). Finally, future work should be directed towards applying and validating the framework on real data.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{setspace}                               
\usepackage[usenames,dvipsnames]{color}  
%This bit is how to use hyperlink without  coloured rectangles around it
\usepackage{float}
\usepackage{hyperref}
\hypersetup{colorlinks,linkcolor=black,filecolor=black,urlcolor=black,citecolor=black}
\usepackage{psfrag}
\newcommand{\dean}[1]{\textcolor{green}{#1}}
\newcommand{\parham}[1]{\textcolor{blue}{#1}}
\newcommand{\vik}[1]{\textcolor{red}{#1}}

@@

\begin{abstract}
%\boldmath
\vik{Neural fields are spatially continuous state variables described by integro-differential equations, which are well suited to describe the spatiotemporal evolution of cortical activations on multiple scales. Here we} develop a multi-resolution approximation (MRA) framework for the integro-difference equation (IDE) neural field model based on semi-orthogonal cardinal B-spline wavelets. State and parameter estimation is performed using the Expectation Maximization (EM) algorithm. A synthetic example is provided
to demonstrate the framework.
\end{abstract} 
\begin{keyword}
  Neural Field Model, multi-resolution approximation (MRA), Expectation Maximization (EM) algorithm
\end{keyword}

@@

Neural field models describe the mass action of the central nervous system and are a critical link in our understanding of the biophysics leading to the generation of  Electroencephalography (EEG) and Magnetoencephalography (MEG) signals. The key components of these models typically describe the macroscopic dynamics of the human brain, but can also be descriptive of finer-scale neurodynamics. Examples are the post-synaptic response kernels (\vik{pulse to wave conversion}) and the activation function (\vik{wave to pulse conversion}) \citep{Jirsa1996,Jirsa1997}. The former can be descriptive of inhibitory and excitatory post-synaptic potentials \vik{in populations of microscopic single neural models, such as integrate and fire neurons,} or an ensemble response to stimulation in macroscopic neural mass models \citep{Stefanescu2008}. The latter, can be used to model spiking statistics across spatial scales from single neurons (where it can be considered a cumulative distribution function (CDF) of the probability of firing) \vik{as} mean firing rates of neural masses. The key factor linking the functions of \vik{neural mass models} to particular spatial scales is \vik{the connectivity}, hence the ability to infer a multi-resolution connectivity kernel will enable the formulation of a multi-resolution neural field model.

The utility of these models for answering questions in clinical neuroscience and neurology lies in the ability to infer parameters from electrophysiological data. The ability to create patient-specific neural models will contribute to our knowledge base of diseases such as epilepsy and will enable the development of new treatment strategies. \vik{Large-scale neuroinformatic efforts such as the Virtual Brain project (\href{http://www.thevirtualbrain.org}{www.thevirtualbrain.org} \citep{Jirsa2010}) are on the way to enable patient-specific full brain modeling.} This is particularly relevant to the advent of new devices that use therapeutic electrical stimulation. Stimulation strategies for devices currently operate in an open loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists enormous potential to improve the performance of these devices using systems theory, which in turn requires a suitable dynamic model. Neural mass and neural field models are ideal candidates for estimation algorithms due to their strong links with physiology and their parsimony. It is expected that parameters of the neural fields models will be patient-specific and will, therefore, need to be inferred from data. \vik{The multi-resolution architecture of the brain and connectivity in particular will play an important role in this context for multiple reasons. First, there is an on-going debate about the organizational level of the brain at which cognition and behaviour is best represented. While clearly single neurons show response changes that map back to learning, memory and motor behaviour, much of the recent work in neuroscience suggest that these elementary features must be organized at a higher level to provide the richness and diversity of the behaving system.  For example, functional neuroimaging work has confirmed that specific parts of visual cortex are more responsive to particular types of stimuli (e.g. faces), but the specificity of this response is dependent on the large-scale pattern of activity across most visual cortical areas \citep{Haxby2001}. Further, the evolution of diffusion magnetic resonance imaging (dMRI) has led to the development of tensor-based tractography, which allows for in vivo images of the white matter skeleton of any brain.  Improvements of the MR sequences (e.g. diffusion spectrum imaging) and tractography analysis have made it possible to use tractography as a standard to define connectivity \citep{Hagmann2008}. Hence individualized connectivity matrices can be developed for brain models enabling patient-specific functional studies and need to be estimated from the experimental data.}

The first work (to the authors best knowledge) describing data-driven mesoscopic neural modeling used a neural mass model to fit EEG data~\citep{Valdes1999}. This approach was extended to coupled neural masses through a Bayesian estimation scheme \vik{called} Dynamic Causal Modeling (DCM)~\citep{David2003}. Following this work, data-driven modeling was extended to continuum field equations that explained the richer dynamics of spatiotemporal neural fields \citep{Galka2008,schiff2008kalman,Daunizeau2009, Pinotsis2012}. Most recently, a framework was developed where a finite element model of the neural field (via a global Galerkin projection) was formed, using a basis function decomposition, to transform the PDE neural field equations into a finite dimension system to facilitate efficient state and parameter estimation~\citep{Freestone2011}. This paper is an extension to the aforementioned study, where we derive a neural field state-space model that accounts for the multi-resolution architecture and spatial dynamics of the human cortex. In this way, a flexible framework is created, whereby both macroscopic and microscopic behavior of the system can be represented simultaneously and completely.

@@

\caption{{\bf Examples of cardinal B-spline functions}. Piecewise polynomial functions (dotted lines) are joined at the break points (black dots). (\textbf{A}) B-spline function of order 2 (linear B-spline). (\textbf{B}) B-spline function of order 8.}

@@

The $m^{th}$  order cardinal B-spline function is defined by the recurrence relation \citep{Chui1992} 
\begin{equation}
N_{m}\left(r\right) = \left(N_{m-1}\ast N_{1}\right)\left(r\right) = \int_0^{1} N_{m-1}\left( r-r'\right)\,\mathrm{d}r',
\label{SplineConvolutionIntegral}
\end{equation}
where $m>1$, $\ast$ denotes convolution, and $N_1\left(r\right)$ is the characteristic function of the unit interval $\left[ 0,1\right)$, also known as indicator function, i.e.
\begin{equation}
N_{1}\left(r\right)=
\begin{cases}
1 & \text{if $ 0\le r<1$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{equation}
Following this, a B-spline of any order, $N_m(r)$, can be computed using the recursive expression defined by \citet{DeBoor2001}
\begin{equation}\label{eq:MRA-DoBoorFormula}
 N_{m}\left(r\right)=\frac{r}{m-1}N_{m-1}\left(r\right)+\frac{m-r}{m-1}N_{m-1}\left(r-1\right) \quad m>1.
 \end{equation}
The most commonly used form of splines is the cubic spline $\left(m=4\right)$, comprised of third degree polynomials added together at the joining points, with an explicit expression given by
\begin{align}
3!N_{4}\left(r\right)=
\begin{cases}
r^3 & \text{if $ 0\le r\le1$}, \\
4-12r+12r^2-3r^3 & \text{if $1\le r\le2$}, \\
-44+60r-24r^2+3r^3 & \text{if $2\le r\le3$}, \\
64-48r+12r^2-r^3 & \text{if $3\le r\le4$}, \\
0 & \mathrm{elsewhere}.
\end{cases}
\end{align}
One important feature of the B-spline functions is the partition of unity property, meaning that the unity can be expressed as a linear sum of B-splines, i.e
\begin{equation}
	\sum_{l}N_m(r+l)\equiv1.
	\end{equation}
This is important as the underlying field can be reconstructed when it is saturated without introducing numerical errors which is not the case for example when Gaussian basis functions are used.

@@

 		\caption{{\bf The scaling and the wavelet function decomposition}. (\textbf{A}) The scaling function (dotted line) at $j=0$ is obtained using 625 scaling functions (solid lines) at $j=4$. (\textbf{B}) The wavelet function (dotted line) at $j=0$ is obtained using 1250 scaling functions (solid lines) at $j=4$.}

@@

\vik{viktor suggested to delete this part: It is important to point out there is no doubt that there will always be a mismatch between the model and actual cortical tissue. Nevertheless,} the simplifications used in formulating the equations enable the development of new methods for extracting information from electrophysiological data. The trade-off between parsimony and complexity is always difficult when considering neurodynamics and this is further complicated when using estimation algorithms that have huge computational demands with such high-dimensional systems.

@@

In this paper, we have presented a novel model-based framework for estimating cortical dynamics from electrophysiological measurements. The novel and key developments of the paper include the multi-resolution finite element model representation of the neural field, and the estimator for an intracortical connectivity kernel with an arbitrary shape. This work is significant, as the ability to create patient-specific neural field models has the potential to contribute to our understanding and improve treatment of diseases resulting from abnormal neurodynamics. Other groups have previously highlighted the importance of a multi-resolution approach in neural field modeling, where the dynamics and connectivity structure differs at different spatial scales~\citep{Jirsa2002,Jirsa2004,Qubbaj2009,Breakspear2005,Schultze-Kraft2011}. 

@@

% In order to apply the framework to real data some assumptions must be made. A critical assumption is that the model provides an apt description of the cortical dynamics.
 \subsection{Modification to higher dimensions}
%To demonstrate MRAIDE framework, an assumption was made where the firing rate behaves linearly. The implementation of the framework in the nonlinear case can be addressed as future work. 
%Another future avenue would be the modification of the estimation framework to perform efficiently at higher dimensions.
 The reconstruction of the neural field in higher dimensions requires $2^n-1$ wavelets, where $n$ is the dimension of the space. For example a two-dimensional MRA can be implemented using 2-D scaling and wavelet functions built up using the tensor-product approach \citep{Meyer1992}. In this case three wavelet functions are required to  extract fine features of the field at vertical, horizontal and diagonal orientations. The formulations provided in this paper can be easily extended to two dimensions. However, the computational load of the estimation algorithm will be significant. It is important to investigate methods for choosing basis functions to balance the complexity with the reconstruction capability of the model.
\subsection{Extensions to the framework} 
Clearly there is, and will always be, a discrepancy between the dynamics of the model and the experimental observation. Nevertheless, the model-based framework proposed in this paper may enable meaningful state tracking and connectivity estimation. The MRAIDE approach is not limited to neural fields; the framework can be applied to modeling other multi-resolution spatiotemporal dynamical systems such as weather systems, ecological systems, and others~\citep{Wikle2002,Xu2005}. The key development is the multi-resolution decomposition forming the state-space model. While the decomposition holds for more sophisticated models, efficiently performing nonlinear smoothing in the large state-space remains a challenge.

Additional extension for future work include extending the model to capture a second order synaptic response kernel, The second-order postsynaptic kernel proposed in \citet{VanRotterdam1982} leads to dynamics that are more reminiscent to real postsynaptic potentials. Another extension of this framework is the inclusion of finite distant-dependent action potential propagation velocity into the parametric neural field equations. \vik{In small networks the resulting delays via propagation can be neglected. However, delays must be accounted for in large scale networks. For instance, \citet{Ghosh2008} demonstrated that inclusion of time delays arising from propagation along connecting fibres are essential for the understanding of the fluctuations in resting state networks (see \citet{Deco2010} for review). In fact, \citet{Ghosh2008} emphasize that the couplings in large-scale brain networks have a space-time structure composed of their anatomical connectivity (space) and signal transmission delays (time). To capture correctly the constraints posed upon the network dynamics by connectivity, the space-time structure of the brain connectivity must be correctly estimated.} Finally, future work should be directed towards applying and validating the framework on real data.

\section{Acknowledgements}
The research reported herein was partly supported by the Brain Network Recovery Group through the James S. McDonnell Foundation and the FP7-ICT BrainScales.
